Inferring true changes in microbial abundance from taxonomically-biased microbiome measurements

### 2021-11-08 appendix-regression.Rmd
Move regression results into appendix

# (APPENDIX) Appendix {-}

# Proofs of regression results

## General regression

Regression analysis can often be framed as seeking the regression function $r(x)$ that describes how the expected value of some response variable $Y$ varies with the a vector of covariates $X$ (@wasserman2004allo),
\begin{align}
  r(x) = E [Y \mid X = x].
\end{align}
For example, the response might be log absolute abundance of a particular taxon.
In our case, however, we don't know $Y$, but rather a measure that is subject to random and systematic error, which I call $Z$.
Let $D = Z - Y$ be the difference between the true response $Y$ and its measurement $Z$, so that $Z = Y + D$.
It follows that
\begin{align}
  (\#eq:regression-error)
  E [Z \mid X] = E [Y \mid X] + E [D \mid X].
\end{align}

## Linear least-squares regression

```{theorem, label = "regression-linearity", name = "Linearity of regression coefficients", echo = TRUE}
Consider a scalar response variable with $J$ observations, which I represent by the vector $y$, that equals a sum of $K$ component response variables $y^{(k)}$ each scaled by a non-zero factor $c^{(k)}$, 
\begin{align}
  (\#eq:y-sum)
  y = \sum_{k = 1}^K c^{(k)} y^{(k)}.
\end{align}
(We may need to assume that the $y^{(k)}$ are linearly independent.)
Let $X$ be a $J$-by-$p$ matrix of covariates for the $J$ observations, with linearly independent columns.
Let $\hat \gamma$ denote the least-squares estimate of the coefficient matrix $\gamma$ in the linear regression
\begin{align}
  y = X \gamma + \epsilon.
\end{align}
Similarly, let $\hat \gamma^{(k)}$ denote the least-squares estimates for the $K$ linear regressions
\begin{align}
  y^{(k)} = X \gamma^{(k)} + \epsilon^{(k)}.
\end{align}
The least-squares coefficient estimates for $y$ are given by the sum of those of the $y^{(k)}$,
\begin{align}
  \hat \gamma = \sum_{k=1}^K c^{(k)} \hat \gamma^{(k)}.
\end{align}
```

```{proof, echo = TRUE}
The Moore-Penrose pseudoinverse of $X$ is $X^+ = (X^T X)^{-1} X^T$.
The least squares estimates are given by multiplying the matrix $X^+$ by the corresponding response vector ([Wikipedia](https://en.wikipedia.org/wiki/Ordinary_least_squares#Matrix/vector_formulation)), so that $\hat \gamma = X^+ y$ and $\hat \gamma^{(k)} = X^+ y^{(k)}$ (for $k = 1, \dots, K$).
The result then follows from the stipulation \@ref(eq:y-sum) and the linearity of matrix multiplication,
\begin{align}
  \hat \gamma 
  = X^+ y 
  = X^+ \sum_{k = 1}^K c^{(k)} y^{(k)}
  = \sum_{k = 1}^K c^{(k)} X^+ y^{(k)}
  = \sum_{k = 1}^K c^{(k)} \hat \gamma^{(k)}.
\end{align}
```

We can use \@ref(thm:regression-linearity) to describe how measurement error, such as that caused by experimental bias, affects least-squares estimates of regression coefficients.

```{theorem, label = "regression-error", echo = TRUE, name = "Error in regression coefficients"}
Suppose that $y$ is the response variable we wish to understand, $z$ is our imperfect measurement of $y$, and $d = z - y$ the difference between the two, so that $z = y + d$.
Let each be a vector of length $J$ describing a set of $J$ observations.
Let $X$ be a $J$-by-$p$ matrix of covariates for the $J$ observations, with linearly independent columns.
Consider the linear regression equations for $y$, $d$, and $z$,
\begin{align}
  y &= X \gamma^{(y)} + \epsilon^{(y)} \\
  z &= X \gamma^{(z)} + \epsilon^{(z)} \\
  d &= X \gamma^{(d)} + \epsilon^{(d)}.
\end{align}
The relationship between the least-squares estimates mirrors that between the variables themselves,
\begin{align}
  (\#eq:regression-coefficient-error)
  \hat \gamma^{(z)} = \hat \gamma^{(y)} + \hat \gamma^{(d)}.
\end{align}
```

```{proof, echo = TRUE}
The result follows directly from $z =  y + d$ and \@ref(thm:regression-linearity).
```

```{corollary, label = "regression-error-simple", echo = TRUE}
For the special case of simple linear regression,
\begin{align}
  y &= \gamma_0^{(y)} + \gamma_1^{(y)} x + \epsilon^{(y)} \\
  z &= \gamma_0^{(z)} + \gamma_1^{(z)} x + \epsilon^{(z)} \\
  d &= \gamma_0^{(d)} + \gamma_1^{(d)} x + \epsilon^{(d)},
\end{align}
the estimates for the intercept and slope coefficients satisfy
\begin{align}
  \hat \gamma_0^{(z)} &= \hat \gamma_0^{(y)} + \hat \gamma_0^{(d)} \\
  \hat \gamma_1^{(z)} &= \hat \gamma_1^{(y)} + \hat \gamma_1^{(d)}.
\end{align}
```

Next, aiming to show the forms for the different transformations of interest.
It might be useful to cover the case where $d$ is split into a constant and a sample-specific part, since probably everything of interest will follow from that.

What, if any general results should go in the main text?

Also relate the expected values of $y$ and $z$ conditional on $X$?.

To figure out what I need:

1. extend the below result for log proportions to multiple regression
2. work out the derivation in terms of \@ref(thm:regression-error)
3. Write a corollary if it seems to be useful for doing all the cases (e.g. logit etc)

### Linear regression of microbiome abundances

First, consider log proportions.
In this case, the intercept coefficient is biased upward by $\log B_i$ and the other coefficients are biased downward by the corresponding coefficient of $\log \bar B$, based on the error formulas for individual samples.

TODO: write the general result (for multiple regression, assuming that an intercept term is present.

```{proposition, echo = TRUE}
Consider the linear equations
\begin{align}
  (\#eq:least-squares-formulation)
  \log a_i    &= \alpha_0 + \alpha_1 x \\
  \log \bar B &= \beta_0 + \beta_1 x   \\
  \log m_i    &= \gamma_0 + \gamma_1 x,
\end{align}
and let $\hat \alpha_0$, etc. denote the least-squares estimates for the regression coefficients supposing perfect information (i.e., that the $a_i$ and $B_i$ are known).
The coefficients for $\log m_i$ are related to those for $\log a_i$ and $\log \bar B$ through the equations
\begin{align}
  (\#eq:least-squares-estimates)
  \hat \gamma_0 &= \hat \alpha_0 + \log B_i - \hat \beta_0 \\
  \hat \gamma_1 &= \hat \alpha_1 - \hat \beta_1.
\end{align}
```

```{proof, echo = TRUE}
```

>>>>

# Error in estimated regression coefficients and standard errors {#appendix-regression}

Results from statistics on regression under measurement error can help us understand the effect of bias on DA regression analysis.

Consider Equation \@ref(eq:regression) for the simple linear regression of log density of a species $i$ on covariate $x$.
Let $y$ stand for the actual log density of a focal species $i$, $z$ stand for the log density that we've measured, and $d = z - y$ equal the difference between the two, which here is the log efficiency of species $i$ minus the log mean efficiency of the sample.
Let $s_{xy}$ denote the sample covariance between variables $x$ and $y$, $s^{2}_{x} = s_{xx}$ and $s_{x}$ denote the sample variance and standard deviation, and $r_{xy} = s_{xy}/(s_{x}s_{y})$ denote the sample correlation.

The ordinary least squares (OLS) and maximum likelihood (MLE) estimates of the slope of $z$ equals the sample covariance of $z$ and $x$ divided by the sample variance in $x$ or, equivalently, the sample correlation of $z$ and $x$ multiplied by the ratio of their sample standard deviations,
\begin{align}
  \hat \beta_z = \frac{s_{zx}}{s^2_x} = r_{zx} \cdot \frac{s_y}{s_x}.
\end{align}
From the (bi)linearity of sample covariances it follows that
\begin{align}
  \hat \beta_z 
  = \frac{s_{yx}}{s^2_x} + \frac{s_{dx}}{s^2_x} 
  = \frac{r_{yx} s_y}{s_x} + \frac{r_{dx} s_d}{s_x} 
  = \hat \beta_y + \hat \beta_d,
\end{align}
where $\hat \beta_y$ and $\hat \beta_d$ denote the slope estimates for $y$ and $d$ (were these values to be known).
The absolute error in the estimate $\hat \beta_{z}$ of $\hat \beta_{y}$ is therefore $\hat \beta_{d}$; it is large in a practical sense when $\hat \beta_{d}$ is large (in absolute value) compared to $\hat \beta_{y}$, which corresponds to the covariance of $d$ with $x$ being large compared to the covariance of $y$ with $x$.

In our case, the covariance of $d$ equals the negative of the covariance of log mean efficiency with $x$.
The absolute error in $\hat \beta$ equals the negative covariance of log mean efficiency scaled by the variance in $x$.
This absolute error is the same for all species; however, its practical significance varies depending on its magnitude relative to that of the slope of the actual log densities.
For species that covary with $x$ more strongly than the log mean efficiency, the error will be relatively small.
This situation might occur either because the mean efficiency varies relatively little across samples or because its variation is relatively less correlated with $x$ compared to the log density of the focal species.

We can similarly understand the impact of measurement error on the precision of our slope estimates.
The OLS and MLE estimated standard error in $\hat \beta$ are both approximately
\begin{align}
  \hat{\text{se}}(\hat \beta)
  \approx \frac{s_{\hat \varepsilon}}{s_x \sqrt{n}},
\end{align}
where $s_{\hat \varepsilon}$ is the sample standard deviation of the residuals
(@wasserman2004allo Chapter 13).
The sample residuals of $z$, $y$, and $d$ have a similar relationship to the regression coefficient estimates,
\begin{align}
  \hat \varepsilon_z 
    &\equiv z - \hat \beta_z x
  \\&= (y + d) - (\hat \beta_y + \hat \beta_d) x
  \\&= (y - \hat \beta_y x) + (d - \hat \beta_d x)
  \\&= \hat \varepsilon_y + \hat \varepsilon_d.
\end{align}
(Note, here I've omitted the subscript indicating the dependence on the sample.)
It follows that the sample variances of the residuals of $z$, $y$, and $d$ are related through
\begin{align}
  s^2_{\hat \varepsilon_{z}} 
  = s^2_{\hat \varepsilon_{y} + \hat \varepsilon_{d}}
  = s^2_{\hat \varepsilon_{y}} + s^2_{\hat \varepsilon_{d}} + 2 s_{\hat \varepsilon_{y} \hat \varepsilon_{d}}.
\end{align}
The standard deviation of the $z$ residuals is increased above that of the $y$ residuals when the $d$ residuals are either uncorrelated or positively correlated with the $y$ residuals, but may be decreased when the $y$ and $d$ residuals are negatively correlated.

In our case, the $d$ residuals equal the negative residuals of the log mean efficiency.
It is plausible that for most species, their residual variation will have a small covariance with log mean efficiency and the net effect of variation in the mean efficiency will be to increase the estimated standard errors, as occurs with most species in Figure \@ref(fig:regression-example).
However, high-efficiency species that vary substantially in proportion across samples may be strongly positively correlated with log mean efficiency such that the estimated standard errors decrease, as we see with Species 9 in Figure \@ref(fig:regression-example).

### 2021-11-08 differential-abundance.Rmd
Fix figure reference
Move regression results into appendix

The bottom row of Figure \@ref(fig:error-proportions) illustrates how variation in mean efficiency leads to error in the inferred fold changes between a pair of samples.
The mean efficiency decreases by a factor of 2.6 (FC: 0.4X) from Sample 1 to Sample 2.
Consequently, the FC of each species is measured to be 2.6X larger than the true value.
Though the fold error for all species is the same, the implications depend on the actual FC and correspond to three distinct types of error: an increase in magnitude, a decrease in magnitude, and a change in direction (or sign).
We can see each type of error in Figure 2.
For Species 1, which increases and thus moves in the opposite direction of the mean efficiency, we see an increase in magnitude of the measured FC (actual FC: 2.3X, measured FC: 6.5X).
For Species 2, which decreases and thus moves in the same direction as the mean efficiency but by a larger factor, we see an decrease in magnitude (actual FC: 0.15X, measured FC: 0.44X).
For Species 2, which decreases by a smaller factor than the mean efficiency, we see a change in direction (actual FC: 0.6X, measured FC: 1.8X), such that the species actually appears to increase!

@@

The variation in measurement error created by the log mean efficiency impacts the point estimate and precision of the regression coefficient $\beta$.
Box \@ref(regression-error) mathematically describes this effect for estimation via Ordinary Least Squares (OLS) or Maximum Likelihood Estimation (MLE), which Figure \@ref(fig:regression-example) illustrates using simulated data.
Variation in the log mean efficiency that is associated with the covariate $x$ creates a systematic error in the estimated slope $\hat \beta$ equal to the negative of the (scaled) covariance of log mean efficiency with $x$.
The absolute error is the same for all species; however, its relative value depends on the magnitude of the covariance of the log mean efficiency with $x$ relative to that of the response (here, $\log \text{density}_{i}$) with $x$ or, equivalently, the relative magnitudes of their slopes.
As in the case of fold changes between pairs of samples, the net effect can be decreases in magnitude (Species 9, 10, and 1 in Figure \@ref(fig:regression-example)), changes in sign (Species 5), or increases in magnitude (remaining species) depending on these relative values.
Variation in the log mean efficiency that is uncorrelated with $x$ does not systematically distort $\hat \beta$ but does affect its precision, typically leading to increased standard errors as the variation in log mean efficiency effectively acts as an additional source of noise in measured abundance (Figure \@ref(fig:regression-example) D).
The exception is for species whose residual variation is strongly positively correlated with that of log mean efficiency (here, Species 9), which can appear to have less random variation and receive standard errors that are too small.
Decreased magnitudes and increased standard errors can both cause associations to be missed that would otherwise have been detected (Species 10 and 1), while increased magnitudes can turn weak or statistically insignificant associations into strong and statistically significant ones (Species 7, 6 and 4).

@@

::: {.rmdnote}

## Box: Error in estimated regression coefficients and standard errors {#regression-error}

Results from statistics on regression under measurement error can help us understand the effect of bias on DA regression analysis.

Consider Equation \@ref(eq:regression) for the simple linear regression of log density of a species $i$ on covariate $x$.
Let $y$ stand for the actual log density of a focal species $i$, $z$ stand for the log density that we've measured, and $d = z - y$ equal the difference between the two, which here is the log efficiency of species $i$ minus the log mean efficiency of the sample.
Let $s_{xy}$ denote the sample covariance between variables $x$ and $y$, $s^{2}_{x} = s_{xx}$ and $s_{x}$ denote the sample variance and standard deviation, and $r_{xy} = s_{xy}/(s_{x}s_{y})$ denote the sample correlation.

The ordinary least squares (OLS) and maximum likelihood (MLE) estimates of the slope of $z$ equals the sample covariance of $z$ and $x$ divided by the sample variance in $x$ or, equivalently, the sample correlation of $z$ and $x$ multiplied by the ratio of their sample standard deviations,
\begin{align}
  \hat \beta_z = \frac{s_{zx}}{s^2_x} = r_{zx} \cdot \frac{s_y}{s_x}.
\end{align}
From the (bi)linearity of sample covariances it follows that
\begin{align}
  \hat \beta_z 
  = \frac{s_{yx}}{s^2_x} + \frac{s_{dx}}{s^2_x} 
  = \frac{r_{yx} s_y}{s_x} + \frac{r_{dx} s_d}{s_x} 
  = \hat \beta_y + \hat \beta_d,
\end{align}
where $\hat \beta_y$ and $\hat \beta_d$ denote the slope estimates for $y$ and $d$ (were these values to be known).
The absolute error in the estimate $\hat \beta_{z}$ of $\hat \beta_{y}$ is therefore $\hat \beta_{d}$; it is large in a practical sense when $\hat \beta_{d}$ is large (in absolute value) compared to $\hat \beta_{y}$, which corresponds to the covariance of $d$ with $x$ being large compared to the covariance of $y$ with $x$.

In our case, the covariance of $d$ equals the negative of the covariance of log mean efficiency with $x$.
The absolute error (i.e., the _bias_ in a statistical sense) in $\hat \beta$ is equals the negative covariance of log mean efficiency scaled by the variance in $x$.
This absolute error is the same for all species; however, its practical significance varies depending on its magnitude relative to that of the slope of the actual log densities.
For species that covary with $x$ more strongly than the log mean efficiency, the error will be relatively small.
This situation might occur either because the mean efficiency varies relatively little across samples or because its variation is relatively less correlated with $x$ compared to the log density of the focal species.

We can similarly understand the impact of measurement error on the precision of our slope estimates.
The OLS and MLE estimated standard error in $\hat \beta$ are both approximately
\begin{align}
  \hat{\text{se}}(\hat \beta)
  \approx \frac{s_{\hat \varepsilon}}{s_x \sqrt{n}},
\end{align}
where $s_{\hat \varepsilon}$ is the sample standard deviation of the residuals
(@wasserman2004allo Chapter 13).
The sample residuals of $z$, $y$, and $d$ have a similar relationship to the regression coefficient estimates,
\begin{align}
  \hat \varepsilon_z 
    &\equiv z - \hat \beta_z x
  \\&= (y + d) - (\hat \beta_y + \hat \beta_d) x
  \\&= (y - \hat \beta_y x) + (d - \hat \beta_d x)
  \\&= \hat \varepsilon_y + \hat \varepsilon_d.
\end{align}
(Note, here I've omitted the subscript indicating the dependence on the sample.)
It follows that the sample variances of the residuals of $z$, $y$, and $d$ are related through
\begin{align}
  s^2_{\hat \varepsilon_{z}} 
  = s^2_{\hat \varepsilon_{y} + \hat \varepsilon_{d}}
  = s^2_{\hat \varepsilon_{y}} + s^2_{\hat \varepsilon_{d}} + 2 s_{\hat \varepsilon_{y} \hat \varepsilon_{d}}.
\end{align}
The standard deviation of the $z$ residuals is increased above that of the $y$ residuals when the $d$ residuals are either uncorrelated or positively correlated with the $y$ residuals, but may be decreased when the $y$ and $d$ residuals are negatively correlated.

In our case, the $d$ residuals equal the negative residuals of the log mean efficiency.
It is plausible that for most species, their residual variation will have a small covariance with log mean efficiency and the net effect of variation in the mean efficiency will be to increase the estimated standard errors, as occurs with most species in Figure \@ref(fig:regression-example).
However, high-efficiency species that vary substantially in proportion across samples may be strongly positively correlated with log mean efficiency such that the estimated standard errors decrease, as we see with Species 9 in Figure \@ref(fig:regression-example).

:::
<!-- </details> -->

>>>>

The bottom row of Figure \@ref(fig:error-proportions) illustrates how variation in mean efficiency leads to error in the inferred fold changes between a pair of samples.
The mean efficiency decreases by a factor of 2.6 (FC: 0.4X) from Sample 1 to Sample 2.
Consequently, the FC of each species is measured to be 2.6X larger than the true value.
Though the fold error for all species is the same, the implications depend on the actual FC and correspond to three distinct types of error: an increase in magnitude, a decrease in magnitude, and a change in direction (or sign).
We can see each type of error in Figure \@ref(fig:error-proportions).
For Species 1, which increases and thus moves in the opposite direction of the mean efficiency, we see an increase in magnitude of the measured FC (actual FC: 2.3X, measured FC: 6.5X).
For Species 2, which decreases and thus moves in the same direction as the mean efficiency but by a larger factor, we see an decrease in magnitude (actual FC: 0.15X, measured FC: 0.44X).
For Species 2, which decreases by a smaller factor than the mean efficiency, we see a change in direction (actual FC: 0.6X, measured FC: 1.8X), such that the species actually appears to increase!

@@

The variation in measurement error created by the log mean efficiency impacts the point estimate and precision of the regression coefficient $\beta$.
Appendix \@ref(appendix-regression) mathematically describes this effect for estimation via Ordinary Least Squares (OLS) or Maximum Likelihood Estimation (MLE), which Figure \@ref(fig:regression-example) illustrates using simulated data.
Variation in the log mean efficiency that is associated with the covariate $x$ creates a systematic error in the estimated slope $\hat \beta$ equal to the negative of the (scaled) covariance of log mean efficiency with $x$.
The absolute error is the same for all species; however, its relative value depends on the magnitude of the covariance of the log mean efficiency with $x$ relative to that of the response (here, $\log \text{density}_{i}$) with $x$ or, equivalently, the relative magnitudes of their slopes.
As in the case of fold changes between pairs of samples, the net effect can be decreases in magnitude (Species 9, 10, and 1 in Figure \@ref(fig:regression-example)), changes in sign (Species 5), or increases in magnitude (remaining species) depending on these relative values.
Variation in the log mean efficiency that is uncorrelated with $x$ does not systematically distort $\hat \beta$ but does affect its precision, typically leading to increased standard errors as the variation in log mean efficiency effectively acts as an additional source of noise in measured abundance (Figure \@ref(fig:regression-example) D).
The exception is for species whose residual variation is strongly positively correlated with that of log mean efficiency (here, Species 9), which can appear to have less random variation and receive standard errors that are too small.
Decreased magnitudes and increased standard errors can both cause associations to be missed that would otherwise have been detected (Species 10 and 1), while increased magnitudes can turn weak or statistically insignificant associations into strong and statistically significant ones (Species 7, 6 and 4).

@@
